---
title: "Applying VDJtrack algorithm to Yellow Fever Virus vaccination data"
author: "M.S."
date: "2023-03-25"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(reshape2)
library(stringdist)
```

Load clonotype annotations, metadata and data from the Yellow Fever Vaccination 
time course (0th day = before, 15th day = after). You can substitute the underlying data with your dataset and clonotype annotations with annotations for 
your clonotypes of interest. If you have several groups of samples, you can 
alternatively use code for DLI/non-DLI comparison in the ``code/`` folder.

```{r message=FALSE}
annot <- read_tsv("example/annotations.txt") 
meta <- read_tsv("example/metadata.txt")
data <- meta %>%
  group_by(donor, time) %>%
  group_modify(~ read_tsv(paste0("example/", .x$file.name)))
```

A little trick - annotate with 1 aminoacid substitution, expand the annotation

```{r}
#Compute distances between strings:
get_distances <- function(aa.seq.1, aa.seq.2, threshold = 1,
                          method = "hamming", ...) {
  stringdistmatrix(unique(aa.seq.1), unique(aa.seq.2), 
                   method = method,
                   useNames = T, ...) %>%
    melt %>%
    filter(value <= threshold) %>%
    rename(aa.seq = Var1, aa.seq.db = Var2, dist = value) %>%
    mutate(aa.seq = as.character(aa.seq), aa.seq.db = as.character(aa.seq.db))
}

#An optimized routine that splits by length and processes in chunks(hamming only):
get_1mm_pairs <- function(aa.seq, aa.seq.db, chunks = 64) {
  d <- tibble(aa.seq = unique(aa.seq)) %>%
    mutate(len = nchar(aa.seq),
           chunk.id = rep(1:chunks, length.out = length(unique(aa.seq))))
  
  db <- tibble(aa.seq.db = unique(aa.seq.db)) %>%
    mutate(len.db = nchar(aa.seq.db)) 
  
  d %>%
    group_by(chunk.id, len) %>%
    group_modify(~ get_distances(.x$aa.seq, db %>% 
                                   filter(len.db == .y$len) %>%
                                   .$aa.seq.db))
}

annot <- annot %>%
  group_by(group) %>%
  group_modify(~get_1mm_pairs(data$cdr3aa, .x$cdr3aa) %>%
                 rename(cdr3aa = aa.seq) %>%
                 select(cdr3aa)) %>%
  ungroup %>%
  select(-chunk.id, -len) 
```

Annotate data by size (singleton/doubleton/...) and according to Yellow Fever 
Virus antigen specificity

```{r}
data <- data %>%
  mutate(quantile = case_when(
    count == 1 ~ "singleton",
    count == 2 ~ "doubleton",
    T ~ "3+"
  )) %>%
  mutate(quantile = factor(quantile, levels = c("singleton",
                                                "doubleton",
                                                "3+"))) %>%
  left_join(annot) %>%
  mutate(group = ifelse(is.na(group), "unknown", group)) %>%
  ungroup
```

Identify clonotypes that appeared (spawned) in `after` dataset but were 
absent in `before` time point

```{r}
# count is the frequency in the 'after' dataset
data %>%
  filter(time == "before") %>%
  select(-time) %>%
  left_join(data %>%
              filter(time == "after") %>%
              select(-time, -count, -quantile, -group) %>%
              mutate(spawned = T)) %>%
  mutate(spawned = !is.na(spawned)) -> data.m
```

Calculate sample sizes and compute sampling statistics

```{r}
data.s0 <- data %>%
  group_by(donor, time, group) %>%
  summarise(diversity = n()) %>%
  dcast(donor + group ~ time) %>%
  rename(div.after = after, div.before = before)

data.s <- data.m %>%
  group_by(donor, group, quantile) %>%
  summarise(spawned = sum(spawned), total = n())

data.vdjtrack <- inner_join(data.s, data.s0)

data.vdjtrack
```

Plot results

```{r}
data.vdjtrack %>%
  ggplot(aes(x = div.after / div.before, 
             y = spawned / total,
             color = quantile)) +
  geom_text(aes(label = donor)) +
  geom_smooth(method = "lm") +
  scale_x_log10("Diversity after / before") +
  ylab("Fraction spawned") +
  scale_color_brewer(palette = "Set1") +
  facet_wrap(~group) +
  theme_classic() +
  theme(aspect = 1, legend.position = "bottom")
```

Compute statistics

```{r}
mdl <- lm(spawned.frac ~ quantile + group + div.ratio,
          data = data.vdjtrack %>%
            mutate(div.ratio = log(div.after / div.before),
                   spawned.frac = log(spawned / total)
            ))
summary(mdl)
```

```{r}
#END
```
